A Formalization of Complete Discrete Valuation Rings and Local Fields

### 2023-12-12 HAL/CPP_paper_v2.tex
review modifications

%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title[A Formalization of Complete DVRs and Local Fields]{A Formalization of Complete Discrete Valuation Rings and Local Fields}

%%%
%%% The "author" command and its associated commands are used to define
%%% the authors and their affiliations.
%%% Of note is the shared affiliation of the first two authors, and the
%%% "authornote" and "authornotemark" commands
%%% used to denote shared contribution to the research.
%\iffalse
%\author{Ben Trovato}
%\authornote{Both authors contributed equally to this research.}
%\email{trovato@corporation.com}
%\orcid{1234-5678-9012}
%\author{G.K.M. Tobin}
%\authornotemark[1]
%\email{webmaster@marysville-ohio.com}
%\affiliation{%
%  \institution{Institute for Clarity in Documentation}
%  \streetaddress{P.O. Box 1212}
%  \city{Dublin}
%  \state{Ohio}
%  \country{USA}
%  \postcode{43017-6221}
%}\fi

% \author[M.~I.~de Frutos Fernández]{María Inés de Frutos Fernández}
\author{María Inés de Frutos Fernández}
\authornote{Both authors contributed equally to this research.}
%\authornotemark[1]
\email{maria.defrutos@uam.es}
\orcid{0000-0002-5085-7446}
\affiliation{%
  \institution{Universidad Aut\'onoma de Madrid}
  \streetaddress{Ciudad Universitaria de Cantoblanco}
   \postcode{28049}
  \city{Madrid}
  \country{Spain}
}

%\author[F.~A.~E.~Nuccio]{Filippo Alberto Edoardo Nuccio Mortarino Majno di Capriglio}
\author{Filippo Alberto Edoardo Nuccio Mortarino Majno di Capriglio}
%\authornote{Both authors contributed equally to this research.}
\authornotemark[1]
\email{filippo.nuccio@univ-st-etienne.fr}
\orcid{0000-0002-5318-9869}
\affiliation{%
	\institution{Université Jean Monnet Saint-Étienne, CNRS, Institut Camille Jordan UMR 5208}
	\streetaddress{23, rue du docteur Paul Michelon}
   \postcode{F-42023}
	\city{Saint-Étienne}
	\country{France}
}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{M.~I.~de Frutos Fernández and F.~A.~E.~Nuccio}

@@

In the setting of a formalization work, \fil{the ambiguities described above concerning the group (with zero) $\Gamma_0$ and the normalization of the valuation call for more attention than in pen-and-paper mathematics\st{allowing for this level of flexibility concerning the group (with zero) $\Gamma_0$ where the valuation takes its values, and about different normalizations, becomes more delicate}}. It is in principle possible to let $\Gamma_0$ vary, including it as a (perhaps implicit) variable: yet one must stipulate the existence of an isomorphism $\Gamma_0\cong\wzmZZ$, adding one more variable, so that we rather find it more convenient to consider only valuations that are $\wzmZZ$-valued, in line with the choice made when defining the adic valuation\href{https://leanprover-community.github.io/mathlib_docs/ring_theory/dedekind_domain/adic_valuation.html}{\extlink} attached to a height-one prime of a Dedekind domain. Even with this choice, the above discussion about different possible normalizations of a $\wzmZZ$-valued valuations, all leading to the same mathematical object, suggests that the focus should be put on \emph{normalized} valuations, rather than general ones. With this in mind, 
we call a valuation ``discrete'' when it is $\wzmZZ$-valued and \emph{also} normalized, encoding this notion in the following class\href{https://github.com/LCFT-Lean/local_fields/blob/76ad487d09babdb0018f394a5634526637ee014a/src/discrete_valuation_ring/basic.lean#L125}{\extlink}:
\begin{lstlisting}[caption={Definition of discrete valuation.}, label={code:def_discrete}]
class is_discrete (v : valuation R $\wzmZZ$) := 
(surj : function.surjective v)
\end{lstlisting}

@@

As far as other systems are concerned, the first formalization of the $p$-adic numbers appeared in the Coq UniMath library in~\cite{padicsCoq}. \fil{Beyond the choice of the theorem prover, the two major differences with our work come from different axiomatization settings: the authors of~\cite{padicsCoq} assume the univalence axiom and work in a constructive setting, replacing the notion of being non-zero with a property of being ``apart from zero''. As a consequence, their construction of the $p$-adic integers $\ZZ_p$ follows a completely different path and involves a carrying operator on the ring $\powerseries[\ZZ]$ and an equivalence relation based on univalent ``paths'' rather than equality. The field $\QQ_p$ is then defined as the Heyting field of fractions of $\ZZ_p$, a construction mimicking the usual field of fractions of an integral domain but in the setting of apartness domains. No treatment of the $p$-adic valuation, of the $p$-adic metric and more generally of topological properties is presented \ibid. From the algebraic point of view, neither the properties of being a DVR, a Dedekind domain, or a local ring, nor algebraic extensions of $\QQ_p$, are approached.}

@@

We also provide a bundled version of this definition, called \code{uniformizer}\href{https://github.com/LCFT-Lean/local_fields/blob/76ad487d09babdb0018f394a5634526637ee014a/src/discrete_valuation_ring/basic.lean#L144}{\extlink}. That is, we provide a structure \code{uniformizer}\relax whose terms consists of two fields: an element $\pi$ of the ring~$R$, together with the proof that $\pi$ is a uniformizer for a given valuation. Since any uniformizer is necessarily a member of the unit ball, we decided to make the field $\pi$ in this definition a term of type \code{vR.integer} (as opposed to type \code{R}).
\begin{lstlisting}
structure uniformizer :=
($\pi$ : vR.integer)
(valuation_eq_neg_one : is_uniformizer vR $\pi$)
\end{lstlisting}
\fil{\st{We use t}T}he proposition \code{is_uniformizer} \fil{is useful} when proving lemmas that apply to any uniformizer element, such as the lemma stating that every uniformizer is nonzero:
\begin{lstlisting}
lemma uniformizer_ne_zero {$\pi$ : R}
  (h$\pi$ : is_uniformizer vR $\pi$) : $\pi \ne 0$ := 
\end{lstlisting}
By contrast, the bundled definition is more convenient to prove results that involve fixing a uniformizer. For example, we use it when proving that every nonzero $r : \ball{K}$ can be factor\fil{ed} as the product of a unit by a power of a fixed uniformizer:
\begin{lstlisting}
lemma pow_uniformizer {r : $K_0$} (hr : r $\ne$ 0)
  ($\pi$ : uniformizer v) :
  $\exists$ n : $\mathbb{N}$, $\exists$ u : $K_0^\times$, r = $\pi$.1^n * u :=
\end{lstlisting}

@@

We make an exception to this rule when implementing mixed and equal characteristic local fields, since we do declare \code{valued} instances for them. This creates a mild diamond for $\padicCompl$ and $\FpXCompl$ but this does not cause trouble in our formalization, thanks to the comparison lemma \code{trivial_extension_eq_valuation}\href{https://github.com/LCFT-Lean/local_fields/blob/76ad487d09babdb0018f394a5634526637ee014a/src/discrete_valuation_ring/trivial_extension.lean#L58}{\extlink} mentioned above. Indeed, in all cases where a the inference system would create problems, the lemma allows us to disambiguate the situation: \fil{for an example, the reader can inspect the proof of} \code{FpX_int_completion.equiv_valuation_subring}\href{https://github.com/LCFT-Lean/local_fields/blob/76ad487d09babdb0018f394a5634526637ee014a/src/eq_characteristic/valuation.lean#L97}{\extlink}, \fil{proving that the unit ball $\FpXComplInt$ is isomorphic to the subring of $\FpXCompl$ whose elements have $(X)$-adic valuation less than or equal to (\code{1:} $\wzmZZ$).}

@@

\paragraph{The \code{normed_field} and the \code{valued} instances}
Recall from~\S\ref{subsec:extensions} that to each discrete valuation $v$ on a field $K$ we can associate a nonarchimedean multiplicative norm $\lvert\cdot\rvert_K$.

\fil{When there exists a \st{If $v$ is the }}preferred discrete valuation \fil{$v$} on $K$, we often register it as a \code{valued} instance:
\begin{lstlisting}
    {K : Type*} [field K] [hv : valued K $\wzmZZ$] 
\end{lstlisting}
In this situation, the norm $\lvert\cdot\rvert_K$ associated to $v$ would be the preferred nontrivial norm on $K$ , up to rescaling, and we would like to record a corresponding \code{normed_field} instance on $K$:
\begin{lstlisting}
    {K : Type*} [normed_field K $\wzmZZ$] 
\end{lstlisting}
Note that the datum of a field is embedded into the definition of the \code{normed_field} class, while the class \code{valued} takes the field structure as an argument. It follows that declaring a \code{normed_field} instance on every field that carries a \code{valued} one leads to a loop in the typeclass inference system\footnote{\lean[4] can detect these kinds of simple loops, so this should not be an issue once our project has been ported to \lean[4].}. We \fil{propose below\st{give}} a concrete example to illustrate this problem, for which we studied
a trace of the instance search using the option \code{trace.class_instances} to track the typeclass inference process.

Suppose that we had declared \code{discretely_normed_field} as an instance instead of as a definition, so that every discretely valued field
\begin{lstlisting}
(K : Type*) [field K] [hv : valued K $\wzmZZ$]
  [is_discrete hv.v]
\end{lstlisting}
would \fil{automatically} inherit a \code{normed_field} instance.
Under this assumption, the typeclass inference system \fil{would} get into an infinite loop when searching for an instance of
%}
\code{valuation_ring hv.v.valuation_subring.to_subring,}\relax 
because one of the first results that Lean tries to apply is \code{valuation_ring.of_field}, \fil{that automatically infers a valuation ring structure on every field. In turn,} this reduces the problem \fil{at hand} to finding an instance of \code{field hv.v.valuation_subring.to_subring.}
This leads to a dead end (since the valuation subring is not a field); however, while trying to find this instance, the inference system finds \code{normed_field.to_field}\relax, and hence it starts to search for an instance of
\code{normed_field hv.v.valuation_subring.to_subring}, for which it will try to apply the definition \code{discretely_normed_field}, which as input requires a \code{field}\relax instance on the valuation subring \fil{which recreates the problem discussed above, risulting in \st{. Hence, }}the system get\fil{\st{s}ting} stuck in an infinite loop.

However, observe that no problem arises when putting both a \code{valued} and a \code{normed_field} instance on a specific field, like $\QQ_p$, because the typeclass inference system is able to find these instances on the specific field, and it will not try to \fil{deduce\st{apply}} them \fil{\st{to find}from other} instances \fil{\st{in}on} other types.

>>>>

%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
%\title[A Formalization of Complete DVRs and Local Fields]{A Formalization of Complete Discrete Valuation Rings and Local Fields}

\title{A Formalization of Complete Discrete Valuation Rings and Local Fields}

%%%
%%% The "author" command and its associated commands are used to define
%%% the authors and their affiliations.
%%% Of note is the shared affiliation of the first two authors, and the
%%% "authornote" and "authornotemark" commands
%%% used to denote shared contribution to the research.

% \author[M.~I.~de Frutos Fernández]{María Inés de Frutos Fernández}
%\author{María Inés de Frutos Fernández}
%\authornote{Both authors contributed equally to this research.}
%\authornotemark[1]
%\email{maria.defrutos@uam.es}
%\orcid{0000-0002-5085-7446}
%\affiliation{%
%  \institution{Universidad Aut\'onoma de Madrid}
%  \streetaddress{Ciudad Universitaria de Cantoblanco}
%   \postcode{28049}
%  \city{Madrid}
%  \country{Spain}
%}

%\author[F.~A.~E.~Nuccio]{Filippo Alberto Edoardo Nuccio Mortarino Majno di Capriglio}
%\author{Filippo Alberto Edoardo Nuccio Mortarino Majno di Capriglio}
%\authornote{Both authors contributed equally to this research.}
%\authornotemark[1]
%\email{filippo.nuccio@univ-st-etienne.fr}
%\orcid{0000-0002-5318-9869}
%\affiliation{%
%	\institution{Université Jean Monnet Saint-Étienne, CNRS, Institut Camille Jordan UMR 5208}
%	\streetaddress{23, rue du docteur Paul Michelon}
%  \postcode{F-42023}
%	\city{Saint-Étienne}
%	\country{France}
%}

\author{María Inés de Frutos-Fernández}
\orcid{0000-0002-5085-7446}
\authornote{Both authors contributed equally to this research.}
\affiliation{%
	\institution{Universidad Aut\'onoma de Madrid}
	\city{Madrid}
	\country{Spain}
}
\email{maria.defrutos@uam.es}

\author{Filippo Alberto Edoardo Nuccio Mortarino Majno di Capriglio}
\orcid{0000-0002-5318-9869}
\authornotemark[1]
\affiliation{%
	\institution{Université Jean Monnet Saint-Étienne}
	\city{Saint-Étienne}
	\country{France}
}
\email{filippo.nuccio@univ-st-etienne.fr}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{M.~I.~de Frutos-Fernández and F.~A.~E.~Nuccio Mortarino Majno di Capriglio}

@@

In the setting of a formalization work, the ambiguities described above concerning the group (with zero) $\Gamma_0$ and the normalization of the valuation call for more attention than in pen-and-paper mathematics. It is in principle possible to let $\Gamma_0$ vary, including it as a (perhaps implicit) variable: yet one must stipulate the existence of an isomorphism $\Gamma_0\cong\wzmZZ$, adding one more variable, so that we rather find it more convenient to consider only valuations that are $\wzmZZ$-valued, in line with the choice made when defining the adic valuation\href{https://leanprover-community.github.io/mathlib_docs/ring_theory/dedekind_domain/adic_valuation.html}{\extlink} attached to a height-one prime of a Dedekind domain. Even with this choice, the above discussion about different possible normalizations of a $\wzmZZ$-valued valuations, all leading to the same mathematical object, suggests that the focus should be put on \emph{normalized} valuations, rather than general ones. With this in mind, 
we call a valuation ``discrete'' when it is $\wzmZZ$-valued and \emph{also} normalized, encoding this notion in the following class\href{https://github.com/LCFT-Lean/local_fields/blob/76ad487d09babdb0018f394a5634526637ee014a/src/discrete_valuation_ring/basic.lean#L125}{\extlink}:
\begin{lstlisting}[caption={Definition of discrete valuation.}, label={code:def_discrete}]
class is_discrete (v : valuation R $\wzmZZ$) := 
(surj : function.surjective v)
\end{lstlisting}

@@

As far as other systems are concerned, the first formalization of the $p$-adic numbers appeared in the Coq UniMath library in~\cite{padicsCoq}. Beyond the choice of the theorem prover, the two major differences with our work come from different axiomatization settings: the authors of~\cite{padicsCoq} assume the univalence axiom and work in a constructive setting, replacing the notion of being non-zero with a property of being ``apart from zero''. As a consequence, their construction of the $p$-adic integers $\ZZ_p$ follows a completely different path and \fil{involves a carrying operator} on the ring $\powerseries[\ZZ]$ and an equivalence relation based on univalent ``paths'' rather than equality. The field $\QQ_p$ is then defined as the Heyting field of fractions of $\ZZ_p$, a construction mimicking the usual field of fractions of an integral domain but in the setting of apartness domains. No treatment of the $p$-adic valuation, of the $p$-adic metric and more generally of topological properties is presented \ibid. From the algebraic point of view, neither the properties of being a DVR, a Dedekind domain, or a local ring, nor algebraic extensions of $\QQ_p$, are approached.

@@

We also provide a bundled version of this definition, called \code{uniformizer}\href{https://github.com/LCFT-Lean/local_fields/blob/76ad487d09babdb0018f394a5634526637ee014a/src/discrete_valuation_ring/basic.lean#L144}{\extlink}. That is, we provide a structure \code{uniformizer}\relax whose terms consists of two fields: an element $\pi$ of the ring~$R$, together with the proof that $\pi$ is a uniformizer for a given valuation. Since any uniformizer is necessarily a member of the unit ball, we decided to make the field $\pi$ in this definition a term of type \code{vR.integer} (as opposed to type \code{R}).
\begin{lstlisting}
structure uniformizer :=
($\pi$ : vR.integer)
(valuation_eq_neg_one : is_uniformizer vR $\pi$)
\end{lstlisting}

The proposition \code{is_uniformizer} is useful when proving lemmas that apply to any uniformizer element, such as the lemma stating that every uniformizer is nonzero:
\begin{lstlisting}
lemma uniformizer_ne_zero {$\pi$ : R}
  (h$\pi$ : is_uniformizer vR $\pi$) : $\pi \ne 0$ := 
\end{lstlisting}
By contrast, the bundled definition is more convenient to prove results that involve fixing a uniformizer. For example, we use it when proving that every nonzero $r : \ball{K}$ can be factored as the product of a unit by a power of a fixed uniformizer:
\begin{lstlisting}
lemma pow_uniformizer {r : $K_0$} (hr : r $\ne$ 0)
  ($\pi$ : uniformizer v) :
  $\exists$ n : $\mathbb{N}$, $\exists$ u : $K_0^\times$, r = $\pi$.1^n * u :=
\end{lstlisting}

@@

We make an exception to this rule when implementing mixed and equal characteristic local fields, since we do declare \code{valued} instances for them. This creates a mild diamond for $\padicCompl$ and $\FpXCompl$ but this does not cause trouble in our formalization, thanks to the comparison lemma \code{trivial_extension_eq_valuation}\href{https://github.com/LCFT-Lean/local_fields/blob/76ad487d09babdb0018f394a5634526637ee014a/src/discrete_valuation_ring/trivial_extension.lean#L58}{\extlink} mentioned above. Indeed, in all cases where a the inference system would create problems, the lemma allows us to disambiguate the situation: for an example, the reader can inspect the proof of \code{FpX_int_completion.equiv_valuation_subring}\href{https://github.com/LCFT-Lean/local_fields/blob/76ad487d09babdb0018f394a5634526637ee014a/src/eq_characteristic/valuation.lean#L97}{\extlink}, proving that the unit ball $\FpXComplInt$ is isomorphic to the subring of $\FpXCompl$ whose elements have $(X)$-adic valuation less than or equal to (\code{1:} $\wzmZZ$).

@@

\paragraph{The \code{normed_field} and the \code{valued} instances}
\mi{Recall from~\S\ref{subsec:extensions}} that to each discrete valuation $v$ on a field $K$ we can associate a nonarchimedean multiplicative norm $\lvert\cdot\rvert_K$.

\fil{When there exists a \st{If $v$ is the }}preferred discrete valuation \fil{$v$} on $K$, we often register it as a \code{valued} instance:
\begin{lstlisting}
    {K : Type*} [field K] [hv : valued K $\wzmZZ$] 
\end{lstlisting}
In this situation, the norm $\lvert\cdot\rvert_K$ associated to $v$ would be the preferred nontrivial norm on $K$ , up to rescaling, and we would like to record a corresponding \code{normed_field} instance on $K$:
\begin{lstlisting}
    {K : Type*} [normed_field K $\wzmZZ$] 
\end{lstlisting}
Note that the datum of a field is embedded into the definition of the \code{normed_field} class, while the class \code{valued} takes the field structure as an argument. It follows that declaring a \code{normed_field} instance on every field that carries a \code{valued} one leads to a loop in the typeclass inference system\footnote{\lean[4] can detect these kinds of simple loops, so this should not be an issue once our project has been ported to \lean[4].}. We \mi{\st{propose} discuss below} a concrete example to illustrate this problem, for which we studied
a trace of the instance search using the option \code{trace.class_instances} to track the typeclass inference process.

Suppose that we had declared \code{discretely_normed_field} as an instance instead of as a definition, so that every discretely valued field
\begin{lstlisting}
(K : Type*) [field K] [hv : valued K $\wzmZZ$]
  [is_discrete hv.v]
\end{lstlisting}
would automatically inherit a \code{normed_field} instance.
Under this assumption, the typeclass inference system would get into an infinite loop when searching for an instance of
%}
\code{valuation_ring hv.v.valuation_subring.to_subring,}\relax 
because one of the first results that Lean tries to apply is \code{valuation_ring.of_field}, that automatically infers a valuation ring structure on every field. In turn, this reduces the problem at hand to finding an instance of \code{field hv.v.valuation_subring.to_subring.}
This leads to a dead end (since the valuation subring is not a field); however, while trying to find this instance, the inference system finds \code{normed_field.to_field}\relax, and hence it starts to search for an instance of
\code{normed_field hv.v.valuation_subring.to_subring}, for which it will try to apply the definition \code{discretely_normed_field}, \mi{\st{which} that} as input requires a \code{field}\relax instance on the valuation subring which recreates the problem discussed above, resulting in the system getting stuck in an infinite loop.

However, observe that no problem arises when putting both a \code{valued} and a \code{normed_field} instance on a specific field, like $\QQ_p$, because the typeclass inference system is able to find these instances on the specific field, and it will not try to \fil{deduce\st{apply}} them \fil{\st{to find}from other} instances on other types.
